### 2023-02-08 materials_methods.tex
Explain indicator func, early stop codons

Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citembe{kosiol_ECM_2007} (codon-triplet-ecm).
It also lets the user provide a codon substitution matrix.
The default FST model (codon-triplet-mg) does not allow substitutions from stop codons \TODO{Juan is this still correct? Do we need to mention the extra error rate.}, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.

@@

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame positions.
%
Specifically, let
$P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)$ represent transition probabilities from a standard codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_0 \cdot X_1 \cdot X_2\right) =
\sum_{Y_0 \cdot Y_1 \cdot Y_2} I(Y_p = y) P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)
\]
%
represent the marginal transition probabilities, where
$p \in \{0, 1, 2\}$ is the position of the descendent nucleotide relative to the ancestral reading frame.
COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

>>>>

Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citembe{kosiol_ECM_2007} (codon-triplet-ecm).
It also lets the user provide a codon substitution matrix.
The default FST model (codon-triplet-mg) does not allow early stop codons on the ancestor sequence, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.

@@

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame positions.
%
Specifically, let
$P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)$ represent transition probabilities from a standard codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_0 \cdot X_1 \cdot X_2\right) =
\sum_{Y_0 \cdot Y_1 \cdot Y_2} I(Y_p = y) P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)
\]
%
represent the marginal transition probabilities, where
$p \in \{0, 1, 2\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function defined
%
$I(y) := \{ 1 \text{ if } y = Y_p\text{, } 0 \text{ otherwise}\}$.
%
COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.
